---
title:  Biol/Math 218
subtitle: Introductory assignment
author: Your Name Here
output:
  rmdformats::downcute:
    self_contained: true
    thumbnails: true
    lightbox: true
    gallery: false
    highlight: tango
    downcute_theme: "chaos"
    toc_depth: 3
knit: (function(input, ...) {
    rmarkdown::render(
      input)
  })
---

This short introductory assignment is meant to give you
an opportunity to play around in RStudio and practice
working with .Rmd files. 

**Do these exercises in order**

Exercise 1: Create a code block

This is a code block:

```{r, example_code_block}

x <- 1
y <- 1

print(x + y)

```

If you run the code, you should get the answer of
1 + 1 = 2

In the space below, write a new code block that outputs the answer of 18 minus 5.
Feel free to copy and paste from the block above; however, the name of all code
blocks in a file need to be unique (i.e., this one can't be "example_code_block").


Exercise 2: Download data

Write a code block below to download the CO2 data from R (it is built in).


Exercise 3: Error checking and fixing (assuming your code blocks above are perfect)

Step 1: Try to "knit" this .Rmd file. If you do it correctly, there should be an
error (or multiple errors). Copy any errors you encounter below as text (not in a code block).

Step 2: The issue is that there are errors is in the code block below. 
Fix those errors and knit it again. Note: It may take multiple tries to figure
out all the errors. Add any new errors to the list in **Step 1**.

```{r, example_code_block}

# Download Palmer Penguins data

library(PalmerPegnuins)

palmerdata <- data(penguins)

# Check names of 

# Figure out data classes
class(penguines$species)
class(penguins$Bill_Length_mm)
```


Step 3: Once you have all the errors corrected, you should have an HTML file. 
Submit that HTML file to Brightspace by 8 am on 2/08/2023. 