---
title:  Biol/Math 218 lesson
subtitle: Normal distribution and Z transformation
author: Your Name Here
output:
  rmdformats::downcute:
    self_contained: true
    thumbnails: true
    lightbox: true
    gallery: false
    highlight: tango
    downcute_theme: "chaos"
    toc_depth: 3
knit: (function(input, ...) {
    rmarkdown::render(
      input)
  })
---

## Part 1: Student's t-distribution

This week we will be discussing how to compare between the means of normally
distributed populations using population parameters (mu and sigma). We call these
tests **parametric tests** 

In real-life studies, however, we rarely know the population standard deviation (sigma).
Therefore, we need to estimate it using the standard error of the mean (SE). This
then changes our normal sampling distribution to a **t-distribution**.

`t = ybar - mu / SE_{ybar}`

The t distribution incorporates sampling error, and therefore is wider than 
a standard normal. Because we need to estimate sigma before calculating t, a 
t-distribution will have a degrees of freedom of `n - 1`.

```{r t_dist_df_5}
curve(dt(x, df = 5), from = -4, to = 4,
      ylab = "Probability density",
      xlab = expression(t[5]))
```

The critical value of a t-distribution is combined 5% area under the tails of 
the distribution (2.5% on both sides). Only 2.5% of one tail is considered for a
one-tailed test.

```{r t_critical_value}
# Calculating the critical value for the previous example
t_crit <- qt(0.05 / 2, df = 5)
print(abs(t_crit))
```

```{r t_area_under_curve}
## Two tailed
x <- curve(dt(x, df = 5), from = -4, to = 4,
      ylab = "Probability density",
      xlab = expression(t[5]))

value1 <- 0.5 * t_crit
value2 <- 0.5 * abs(t_crit)

polygon(c(x$x[x$x <= value1], value1),
        c(x$y[x$x <= value1], 0),
        col = "#69b3a2",
        border = 1)
text(x = -3, y = 0.1, labels = "2.5%")

polygon(c(x$x[x$x >= value2], value2),
        c(x$y[x$x >= value2], 0),
        col = "#69b3a2",
        border = 1)
text(x = 3, y = 0.1, labels = "2.5%")

# One tailed

x <- curve(dt(x, df = 5), from = -4, to = 4,
      ylab = "Probability density",
      xlab = expression(t[5]))


polygon(c(x$x[x$x >= value2], value2),
        c(x$y[x$x >= value2], 0),
        col = "#69b3a2",
        border = 1)
text(x = 3, y = 0.1, labels = "2.5%")
```

We can calculate the **95% CI for the mean** using the t critical value

`ybar - t_{0.05(2),df} * SE_{ybar} < mu < ybar + t_{0.05(2),df} * SE_{ybar}`

## Practice problem 1: 95% CI for the mean of *Iris setosa* sepal widths

Let's calculate the 95% CI for the sepal widths of *Iris setosa* from the `iris` 
data set.

```{r 95_ci_setosa}

```


## Part II: T-tests

We can use the t-distribution to perform hypothesis tests related to the means
of populations.

### One-sample t-test

**One-sample t-test**: compare the mean of a sample `m_{1}` to a proposed population mean
`m_{0}`

Hypotheses: 

`H_{0}: mu_{1} = mu_{0}`

`H_{A}: mu_{1} != mu_{0}`

One sample t-statistic:

`t = (ybar - mu_{0}) / SE_{ybar}`

We can then use the t-statistic to calculate a P value in R using the t-distribution

**Assumptions:**

- Data is a random sample from the population
- Y is normally distributed in the population

### Example: Is *Iris veriscolor* petal width equal to the mean petal widths for Iris

**Hypotheses:**

`H_{0}: mu_{1} = mu_{0}`

`H_{A}: mu_{1} != mu_{0}`

```{r one_sample_t}
sample <- iris$Petal.Width[iris$Species == "versicolor"]

ybar <- mean(sample)
n <- length(sample)
mu <- mean(iris$Petal.Width)
SE_ybar <- sd(sample) / sqrt(n)

t_stat <- (ybar - mu) / SE_ybar
print(t_stat)

t_crit <- qt(0.05 / 2, df = n - 1)
print(abs(t_crit))

# Two-tailed P value
2 * pt(q = t_stat, df = n - 1, lower.tail = FALSE)

# 95% CI
lower_ci <- ybar - (abs(t_crit) * SE_ybar)
upper_ci <- ybar + (abs(t_crit) * SE_ybar)

print(paste(round(lower_ci, 3), "< μ <", round(upper_ci, 3)))

# Check work with t.test()
t.test(sample, mu = mu)
```


### Paired t-test

**Paired t-test**: compare the mean difference `m_{d]` between two samples where samples 
are paired  with each other (e.g., before and after observations 
from the each individual in a sample) 

Hypotheses: 

`H_{0}:mu_{d} = 0`

`H_{A}: mu_{d} != 0`

The estimate for `mu_{d}` is `dbar`

Paired t-statistic:

`t = (dbar - mu_{d0}) / SE_{dbar}`

**Assumptions:**

- Data is a random sample from the population
- Paired differences normally distributed in the population


### Example: Weight of squirrels

Weight of squirrels before treatment: 450, 400, 440, 390, 400, 432, 415, 401
Weight of squirrels after treatment: 510, 590, 520, 550, 585, 500, 590, 577

**Hypotheses**

`H_{0}:mu_{d} = 0`

`H_{A}: mu_{d} != 0`

```{r paired_t, warning = FALSE}

before <- c(450, 400, 440, 390, 400, 432, 415, 401)
after <- c(510, 590, 520, 550, 585, 500, 590, 577)

# Produce a plot of the pairs
library(ggpubr)
pairs <- data.frame(before = before, after = after)
ggpaired(pairs, cond1 = "before", cond2 = "after", ylab = "Weight (g)")

# t-statistic
d <- pairs$before - pairs$after
dbar <- mean(d)
d_sd <- sd(d)
n <- length(d)

mu <- 0
SE_dbar <- d_sd / sqrt(n)

t_pairs <- (dbar - mu) / SE_dbar
print(t_pairs)

t_crit <- qt(0.05 / 2, df = n - 1)
print(abs(t_crit))

# Two-tailed P value
2 * pt(q = abs(t_pairs), df = n - 1, lower.tail = FALSE)

# Check work using t.test() function
t.test(before, after, paired = TRUE)
```

## Two sample t-test

We can also compare the means of two samples that are not paired, most comparisons
of means will fall under the **two-sample t-test**.

Hypotheses: 

`H_{0}:mu_{1} = mu_{2}`

`H_{A}: mu_{1} != mu_{2}`

Two sample t-statistic:

`t = (ybar_{1} = ybar_{2}) / SE_{ybar_{1} - ybar_{2}}`

SE calculated as:

`SE_{ybar_{1} - ybar_{2}} = sqrt(s^2_{p} * (1 / (n_{1}) + (1 / n_{2}))`

Where:

`s^2_{p} = (df_{1} * s_{1}^2 + df_{2} * s_{2}^2) / (df_{1} + df_{2})`

This called the **pooled sample variance**

**Assumptions:**

- Data is a random sample from the population
- Data is normally distributed in the population
- Variance between samples in equal

### Example: Penguin bills

Lets test whether there is a difference in bill length between Adelie and Gentoo
penguins.

Hypotheses: 

`H_{0}:mu_{1} = mu_{2}`

`H_{A}: mu_{1} != mu_{2}`

```{r two_sample_t}

library(palmerpenguins)
data(penguins)

penguins <- na.omit(penguins)

adelie_bill <- penguins$bill_length_mm[penguins$species == "Adelie"]
gentoo_bill <- penguins$bill_length_mm[penguins$species == "Gentoo"]


# Visualize data (and check normality)
hist(adelie_bill, xlab = "Bill length",
      main = substitute(paste(italic('Adalie'),  "penguins")),
      col = "#69b3a2")
hist(gentoo_bill, xlab = "Bill length",
      main = substitute(paste(italic('Gentoo'),  "penguins")),
      col = "#b37d69")

# Two-sample t-statistic
ybar_1 <- mean(adelie_bill)
var_1 <- var(adelie_bill)
n_1 <- length(adelie_bill)
df_1 <- n_1 - 1

ybar_2 <- mean(gentoo_bill)
var_2 <- var(gentoo_bill)
n_2 <- length(gentoo_bill)
df_2 <- n_2 - 1

pooled_var <- ((df_1 * var_1) + (df_2 * var_2)) / (df_1 + df_2)

SE_diff <- sqrt(pooled_var * ((1 / n_1) + (1 / n_2)))

t_diff <- (ybar_1 - ybar_2) / SE_diff
print(t_diff)

t_crit <- qt(0.05 / 2, df = n_1 + n_2 - 2)
print(abs(t_crit))

# Two-tailed P value
2 * pt(q = abs(t_diff), df = n_1 + n_2 - 2, lower.tail = FALSE)

# Check work using t.test() function
t.test(adelie_bill, gentoo_bill, var.equal = TRUE)

# If variances are not equal,
# we can use Welch's t-test (which is the default in R)
t.test(adelie_bill, gentoo_bill)
```

## Part III: Checking Assumptions

### Checking normality

Checking normality is relatively straight forward. We can use `hist()` to produce
a histogram of our observations and use the **eyeball test**. If it looks normal
(i.e., like a bell curve), then we can assume our data is normal.

```{r eyeball_test}
library(palmerpenguins)
data(penguins)

penguins <- na.omit(penguins)
adelie_bill <- penguins$bill_length_mm[penguins$species == "Adelie"]

# Visualize data (and check normality)
hist(adelie_bill, xlab = "Bill length",
      main = substitute(paste(italic('Adelie'),  "penguins")),
      col = "#69b3a2")
```

There is also a hypothesis test called the **Shapiro-Wilk test**, which is a 
goodness-of-fit test for the normal distribution. A P-value > 0.05 indicates
that the data is likely to be from a population that is normally distributed.

Hypotheses:

`H_{0} = Data are sampled from a normal distribution`

`H_{A} = Data are not sampled from a normal distribution`

```{r shapiro_test}
library(palmerpenguins)
data(penguins)

penguins <- na.omit(penguins)
adelie_bill <- penguins$bill_length_mm[penguins$species == "Adelie"]

shapiro.test(adelie_bill)
```

### Checking homogeneity of variances

Notice above with the two-sample t-test that we set `var.equal = TRUE`. 
While we can use the "eyeball test" to 
determine if our data is from a normal distribution. We can't eyeball equal
variances. Instead, we can use a hypothesis tests. One way is the **F-test**,
the other is **Levene's test**.

## F-test

Hypotheses:

`H_{0}: var_{1} = var{2}`

`H_{A}: var_{1} != var{2}`

Equation: 

`F = s^2_{1} / s^2_{2}`

Note: Make sure the numerator is the highest of the two variances

```{r f_test}
library(palmerpenguins)
data(penguins)

penguins <- na.omit(penguins)

adelie_bill <- penguins$bill_length_mm[penguins$species == "Adelie"]
gentoo_bill <- penguins$bill_length_mm[penguins$species == "Gentoo"]

var_1 <- var(adelie_bill)
n_1 <- length(adelie_bill)
df_1 <- n_1 - 1

var_2 <- var(gentoo_bill)
n_2 <- length(gentoo_bill)
df_2 <- n_2 - 1

f_stat <- var_1 / var_2
print(f_stat)

# Check work using var.test()
var.test(adelie_bill, gentoo_bill)
```


## Levene's Test

This one is a bit more complicated and allows us to compare between more than two
levels of a group. Here we can check homogeneity of variances between all 
three species of penguins.

Hypotheses:

`H_{0}: var_{1} = var{2}`

`H_{A}: var_{1} != var{2}`

```{r levenes_test}
library(car)
library(palmerpenguins)
data(penguins)

penguins <- na.omit(penguins)

leveneTest(bill_length_mm ~ species, penguins)
```

