---
title:  Biol/Math 218 Portfolio 2
subtitle: Data manipulation and basic plotting
author: Your Name Here
output:
  rmdformats::downcute:
    self_contained: true
    thumbnails: true
    lightbox: true
    gallery: false
    highlight: tangoS
    downcute_theme: "chaos"
    toc_depth: 3
knit: (function(input, ...) {
    rmarkdown::render(
      input,
      output_dir = "../knit_html"
    )
  })
---

## Introduction

In this portfolio, you will demonstrate your skills at uploading data and doing basic 
data manipulation and plotting. Answers to each **question** should be entered in the 
designated `Answer:` block. All **exercises** need to be done within `r{}` chunks, with any 
non-code commented out with `#`. Use the `print()` function to show results of calculations.

- Total exercises: **07**
- Total questions: **06**

## Part 01: Dealing with data

### Exercise 01
In the space below, write a `r{}` code chunk to
  
  1) Call the "iris" data set, without the whole data set being printed
  2) List the column names of the data set
  3) Determine which columns are numeric and which are factors
  4) Determine the levels of the factors

### Exercise 02
In the space below, write a `r{}` code chunk to:
  
  1) Export the `iris` data set as a *.csv file
  2) Import that *.csv as a data frame called "iris_import"
  3) Create a vector of only petal width observations
  4) Create a separate data frame of only the *Iris setosa* observations
  5) Create a separate vector of only *Iris veriscolor* sepal lengths


## Part 02: Basic plotting

**Question 01**: What are the four rules for drawing a good graph?

`Q01 Answer:`

### Exercise 03-01
Run the following code to produce a bar graph
```{r, bar_graphs}

max_temp <- c(22, 27, 26, 24, 23, 26, 28)

## Barplot() function
barplot(max_temp,
        main = "Maximum Temperatures in a Week",
        ylab = "Degree Celsius",
        xlab = "Day",
        ylim = c(0, 30),
        names.arg = c("Sun", "Mon", "Tue", "Wed", "Thu", "Fri", "Sat"),
        col = "darkred",
        horiz = FALSE)

```

**Question 02**: For what sort of data are bar graphs used?

`Q02 Answer:`

**Question 03**: Which day had the highest maximum temperature during that week?

`Q03 Answer:`

### Exercise 03-02
In a new `r{}` code chunk below, recreate this bar graph for the following temperatures: 
31, 27, 24, 23, 25, 28, 22 and make the bars **light blue**

*Hint: Use copy/paste and make sure to change the ylim() to an appropriate range* 

### Exercise 04-01 
Run the following code to plot a scatter plot of *Iris* petal width and petal length
```{r, iris_scatterplot}

data(iris)

plot(iris$Petal.Length, iris$Petal.Width, pch = 16, col = "black", 
     main = "Relationship of flower part dimensions", xlab = "Petal length", 
     ylab = "Petal width",
     xlim = c(4, 7), ylim = c(0, 3))
```

**Question 04**: What is the relationship between petal length and petal width?

`Q04 Answer:`

### Exercise 04-02 
To look at the strength of the relationship, you can 
plot the line of best fit by using the lm() and abline() functions. 
Practice that here, but we will look into this in more depth later 
on in the semester.

```{r, iris_bestfit}
data(iris)

plot(iris$Petal.Length, iris$Petal.Width, pch = 16, col = "black",
     main = "Relationship of flower part dimensions", xlab = "Petal length",
     ylab = "Petal width",
     xlim = c(4, 7), ylim = c(0, 3))

bestfit <- lm(iris$Petal.Width ~ iris$Petal.Length) # Form y ~ x
abline(bestfit, lwd = 3) # lwd: line width
```

### Exercise 04-03
Recreate this plot and line of best fit for the sepal 
length and sepal widths of *Iris* in a code chunk below. 
Change colors if you would like, but make sure to change the x and y limits if needed

### Exercise 05
Run this box and whisker plot and answer the question.
```{r iris_boxplot}
data(iris)

boxplot(iris$Sepal.Length ~ iris$Species,
      main = "Length of Iris sepals",
      ylab = "Sepal length",
      xlab = "Species")
```

**Question 05**: What do the thick lines in the boxes represent? What do the whiskers represent?

### Exercise 06
Run this violin plot and answer the question.
```{r iris_violin}
data(iris)
library(ggplot2)

ggplot(iris, aes(x = Species, y = Sepal.Length, fill = Species)) +
  ggtitle("Length of Iris sepals") +
  xlab("Species") +
  ylab("Sepal length") +
  geom_violin()
```

**Question 06**: Why is this called a violin plot? What do they represent?

## Part 03: Advanced plots

### Exercise 07
Go to https://r-graph-gallery.com/ and:

  1) Choose two plot types that we have not gone over here
  2) Copy and paste code to produce each plot in their own chunk, making sure that they work properly