---
output:
  word_document:
    reference_docx: "../templates/word_template.docx"
---

# Biostatistics (Biol/Math 218) Portfolio 3 
# Summary statistics and data exploration

**Name:**

Answer all questions to the best of your ability. All r code needs to 
be within properly labeled code block and all code needs to work properly, 
so double check your work!

## Exercise 01: Basic mathematical functions

Consider this data: 11, 12, 13, 14, 14, 13, 12, 11

In a code chunk below, produce a vector of the above data and determine these using the built-in functions in r
  1) Number of observations
  2) Sum of observations
  3) Highest observation
  4) Lowest observation


### Exercise 02: In a code chunk, calculate the mean of Iris sp. sepal widths both using the mean equation and using the mean() function.

## Question 01: How do your results from each calculation compare?

## Exercise 03: In a code chunk, calculate the mean of Iris setosa sepal widths


## Exercise 04: Write a code chunk to calculate:
  1) Range of Iris setosa sepal widths
  2) Median of the Iris setosa sepal widths

## Question XX: Is there a skew to the Iris setosa sepal width data? If so, what direction and explain your reasoning

## Exercise XX: Write a code chunk to calculate the variance of the Iris versicolor petal lengths using the var() function.

## Exercise XX: Write a code chunk to calculate the standard deviations of Iris virginica petal lengths using the standard deviation equation and using the sd() function.

## Exercise XX: Write a chode chunk using summary() for the entire Iris dataset. 


### Question XX: What are histograms used to illustrate?

## Exercise XX: Run the following code to produce a histogram

```{r, sepal_histogram}
data(iris) # Use the built-in Iris data set

hist(iris$Sepal.Length, col = "lightblue", main = "Iris sepal length",
        xlab = "Sepal length")
```

## Exercise XX: In a new code chunk below, recreate this histogram for Iris petal width and change the color.

Fun code: Run this chunk to see the built-in colors in R
```{r, color_grid}
pdf()
showCols <- function(cl = colors(), bg = "grey",
                     cex = 0.75, rot = 30) {
    m <- ceiling(sqrt(n <- length(cl)))
    length(cl) <- m * m; cm <- matrix(cl, m)
    require("grid")
    grid.newpage(); vp <- viewport(w = .92, h = .92)
    grid.rect(gp = gpar(fill = bg))
    grid.text(cm, x = col(cm) / m, y = rev(row(cm))/m, rot = rot,
              vp = vp, gp = gpar(cex = cex, col = cm))
  }
showCols(cl = colors(), bg= "gray33", rot = 30, cex = 0.75)
dev.off()
```

## Question XX: Run each of the three code chunks below, identify the shape of each.

```{r, histogram_1}
hist(rbeta(10000,5,2), main = "", xlab = "", col = "darkred")
```

```{r, histogram_2}
mu1 <- log(1)
mu2 <- log(50)
sig1 <- log(3)
sig2 <- log(3)
cpct <- 0.4

histDistFunc <- function (n,cpct, mu1, mu2, sig1, sig2) {
  y0 <- rlnorm(n, mean = mu1, sd = sig1)
  y1 <- rlnorm(n, mean = mu2, sd = sig2)

  flag <- rbinom(n,size = 1,prob = cpct)
  y <- y0 * (1 - flag) + y1 * flag 
}

histData <- histDistFunc(n = 100, cpct, mu1, mu2, sig1, sig2)
hist(log(histData), main = "", col = "darkred")
```

```{r, histogram_3}

hist(rbeta(10000, 1, 1), main = "", xlab = "", col = "darkred")
```