---
title:  Biol/Math 218 lesson
subtitle: Binomial distribution and the binomial test
author: Your Name Here
output:
  rmdformats::downcute:
    self_contained: true
    thumbnails: true
    lightbox: true
    gallery: false
    highlight: tango
    downcute_theme: "chaos"
    toc_depth: 3
knit: (function(input, ...) {
    rmarkdown::render(
      input)
  })
---

This week we will be learning about proportions, the binomial distribution
and the binomial test.

## Part 1: Binomial distribution

**Binomial distribution**: the distribution “successes” of many independent trials. 
There are only two possible outcomes, success and non-success

**Assumptions**

- Each trial is independent of the other
- The number of “n” trials is fixed
- The probability of success, “p”, is the same for each trial

`Pr[X successes] = (n X) * p * (1 - p)^n - x`

`(n X) = n! / (X! * n - X)!`


The proportion of `X` successes out of `n` trials:

`phat = X / n`

The population standard error:

`sigma_phat = sqrt(p(1-p) / n)`

95% confidence interval:

`p' = X + 2 / n + 4 `

`p' - 1.96 * sqrt((p' * (1 - p')) / (n + 4))`

`p' + 1.96 * sqrt((p' * (1 - p')) / (n + 4))`

## Part 2: Binomial test

We can use the binomial distribution to test whether the number of successes 
observed over a number of independent trials is different than what is expected
from the null. We combine the probability getting the number of successes that
we observed as well as getting more successes than we 

### Example: Binomial test

Probability of >= 22 successes
n = 24

`H_0: p = 0.5`

`H_A: p != 0.5`


```{r binom_test_example}

pr22 <- factorial(24) / (factorial(22) * factorial(24 - 22))
x <- pr22 * 0.5^22 * 0.5^(24 - 22)

pr23 <- factorial(24) / (factorial(23) * factorial(24 - 23))
y <- pr23 * 0.5^23 * 0.5^(24 - 23)

pr24 <- factorial(24) / (factorial(24) * factorial(24 - 24))
z <- pr24 * 0.5^24 * 0.5^(24 - 24)

p <- 2 * (x + y + z)

print(p)

phat <- 22 / 24

print(phat)

# 95% CI interval

p_prime <- (22 + 2) / (24 + 4)

CI_minus <- p_prime - 1.96 * sqrt((p_prime * (1 - p_prime)) / (24 + 4))
CI_plus <- p_prime + 1.96 * sqrt((p_prime * (1 - p_prime)) / (24 + 4))

print(paste(CI_minus, "±", CI_plus, sep = " "))

# Built-in R function
binom.test(x = 22, n = 24, p = 0.5)
```


Note: The binomial test calculates an **exact** P-value

### Practice problem 1: Binomial test, bird care


In some species of birds, juveniles help other adult birds to raise offspring. 
Scientists are unsure what the evolutionarily advantage of this sort of altruism 
is in the wild, however, it is postulated that these juveniles are likely to be 
preferentially helping adults that they are more closely related to (kin selection).
To test this, 20 birds that each lived nearby two nests of adults, one with relatives,
the other with non-relatives. It was observed that 16 of the birds helped their
kin, while 4 helped the non-kin. 

1. Is there preferential care? 
2. What is the probability of success, including the 95% CI?

`Hypotheses`

```{r bird_binomial}
```


