---
title:  Biol/Math 218 Portfolio 5
subtitle: Probability
author: Your Name Here
output:
  rmdformats::downcute:
    self_contained: true
    thumbnails: true
    lightbox: true
    gallery: false
    highlight: tango
    downcute_theme: "chaos"
    toc_depth: 3
knit: (function(input, ...) {
    rmarkdown::render(
      input,
      output_dir = "knit_portfolios"
    )
  })
---

## Introduction 
In this portfolio, you will demonstrate your skills describing probability rules. 
Answers to each **question** should be entered in the 
designated `Answer:` block. All **exercises** need to be done within `r{}` chunks, with any 
non-code commented out with `#`. Use the `print()` function to show results of calculations.


- Total exercises: **04**
- Total questions: **06**

## Part 01: Mutually exclusive vs. non-exclusive events

Imagine that you have a population of 100 cats, 40 that are black with long fur,
30 that are white
with short fur, 10 that are black with short fur, and 20 that are white with
long fur.

**Question 01**: Are "black" and "long fur" mutually exclusive traits for this
population of cats?

`Q01 Answer: No they are not`

**Question 02**: Are "white" and "short fur" independent traits for this population of cats?

`Q02 Answer: Yes they are`

## Part 02: Probability basics

### Exercise 01
A normal deck of cards has 52 cards, consisting of 13 each of four suits:
spades, hearts, diamonds, and clubs. Hearts and diamonds are red, while
spades and clubs are black. Each suit has an ace, nine cards numbered 2
through 10, and three face cards. The face cards are a jack, a queen, and a king.
Answer the following questions for a single card drawn at random from a
well-shuffled deck of cards.

Answer the following `five` questions in a code chunks, show all calculations

Make sure to write out the probability equation first
Example: `Pr[Queen or Ace] = Pr[Queen] + Pr[Ace]`

1) What is the probability of drawing a king of any suit? `4/52 or 1/13 or 0.077`
2) What is the probability of drawing a face card that is also a spade? `3/52 or 0.058`
3) What is the probability of drawing a card without a number on it? `16/52 or 4/13 or 0.31`
4) What is the probability of drawing an ace? `4/52 or 1/13 or 0.077`
5) What is the probability of drawing a red ace? `2/52 or 1/26 or 0.038`

**Question 03**: Are these events ("ace" and "red") mutually exclusive? Are they independent?

`Q03 Answer:`

**Question 04**: List two possible events that are mutually exclusive for a single draw from a deck of cards.

`Q04 Answer: No, Yes`

## Part 03 General addition principle

### Exercise 02
When asked about their thoughts on the movie the Batman,
20% of movie-goers said that it was "excellent",
10% said that it was "pretty good"," 53% were "indifferent,"
16% said that "it was pretty bad" and 1% said it was "especially horrible".
Only one answer per movie-goer was allowed.

**Question 05**: Are these five possible answers mutually exclusive? Explain.

`Q05 Answer: Yes because each movie-goer was only allowed one answer`

Answer the following `two` questions in a code chunks, show all calculations
Make sure to write out the probability equation first

1) What is the probability that a movie-goer had a positive review of this movie?

`## Answer: Pr[E or PR] = Pr[E] + Pr[PR]`
`pr.vr.sr <- 20/100 + 10/100 # 0.3`

2) What is the probability that a movie-goer would review the movie as anything other than especially horrible?

`## Answer: PR[all but EH]
all.but.ed <- 1 - 1/100 # 0.99`
``
## Part 04 General multiplication principle

### Exercise 03
Of a population of males, 60% were smokers. Suppose that men that smoke have
a probability 20.7% of developing lung cancer during their lifetime. In contrast, men who
did not smoke had a 1.3% lifetime chance of developing lung cancer.

Answer the following `five` questions in code chunks, show all calculations
Make sure to write out the probability equation first

1) What is the conditional probability of of a man from this population getting cancer, given that they smoked?

`Answer: Pr[Cancer|Smoked] = 20.7/100 or 0.207`

2) Calculate the probability that a member of this population both smoked and eventually developed lung cancer.

`Answer: pr.c.s <- 60/100 * (20.7/100) # 0.124`

3) Using the general multiplication rule, calculate the probability that a member of this population both did not smoke and never developed lung cancer.

`Answer: pr.ns.nc <- (1 - 60/100) * (1- 1.3/100) # 0.395`

## Part 05 Bayes' Theorem

### Exercise 04
Using the information from `Part 04`, calculate the probability that a man from the 
population above smoked, given that he had been diagnosed with lung cancer in a 
code chunk below.

**Question 06**: What is the Bayes' theorem equation for the case described in this question?
No need to put it in a code chunk.

`Q06 Answer: Pr[Smoked|Cancer] = Pr[Cancer|Smoked] * Pr[Smoked] / Pr[Cancer] `

Now, each in their own code chunk

1) Calculate the probability that a man in this population would develop lung cancer in his lifetime.

```{r}
prc <- ((20.7/100)*(60/100)) + ((1.3/100)*(1 - 60/100))
print(prc) # 0.1294

```

(Use the law of total probability.)
1) Use Bayes' theorem to calculate the probability that a man from this population 
smoked, given that he eventually developed lung cancer.

```{r}
## Answer:  Pr[Smoked|Cancer] = Pr[Cancer|Smoked] * Pr[Smoked] / Pr[Cancer] 
##          Pr[Smoked|Cancer] = (19.1/100) * (58/100) / Pr[Cancer] 
##          Pr[Cancer] = (19.1/100)*(58/100) + (1.6/100)*(1 - 58/100)
pr.s.c <- (20.7/100) * (60/100) / ((20.7/100)*(60/100) + (1.3/100)*(1 - 60/100))
print(pr.s.c) # 0.9598145
```