---
title:  Biol/Math 218 lesson
subtitle: Displaying data
author: Your Name Here
output:
  rmdformats::downcute:
    self_contained: true
    thumbnails: true
    lightbox: true
    gallery: false
    highlight: tango
    downcute_theme: "chaos"
    toc_depth: 3
knit: (function(input, ...) {
    rmarkdown::render(
      input)
  })
---

## Part 1: Figures in R

Today we will be looking at the basics of displaying data using plots and 
tables using R. 

The lesson can be found here: [Presenting Data](https://www.lundquistecology.com/coures/biostatistics/presenting-data)

First we will look at some basic plots. The code for the plots
are located in the [/basic_plotting](/basic_plotting) folder.

### Histograms

```{r histogram}
# Histograms in R
# Call the data
data(iris)

# Built-in plot() function
x <- iris$Sepal.Length


hist(x, col = "#69b3a2", xlab = "Sepal Length",
    main = "Iris sepal lengths")
```

### Scatter plot

```{r scatter}
# Scatter plots in R
# Call the data
data(iris)

# Built-in plot() function
x <- iris$Sepal.Length
y <- iris$Sepal.Width

# R will automatically detect the data type and produce
# a scatter plot
plot(x, y, pch = 16, col = "#69b3a2", xlab = "Sepal Length",
     ylab = "Sepal Width", main = "Iris sepal length vs. width",
     xlim = c(4, 8), ylim = c(1, 5))
```

### Bar plot

```{r bar_plot}
# Bar plot using ggplot2 library
values <- c(60, 20, 30, 20)
condition <- c("Treatment", "Control", "Treatment", "Control")
group <- c("Group 1", "Group 1", "Group 2", "Group 2")

data <- data.frame(values, condition, group)

library(ggplot2)

# Stacked bars
ggplot(data, aes(x = condition, y = values, fill = group)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = c("#69b3a2", "#b37d69"))

# Bars side by side
ggplot(data, aes(x = condition, y = values)) +
  geom_bar(aes(fill = group), stat = "identity", position = "dodge") +
  scale_fill_manual(values = c("#69b3a2", "#b37d69"))
```

### Violin plot

```{r violin}
# Violin plots using ggplot2 library
data(iris)

ggplot(iris, aes(x = Species, y = Sepal.Length, fill = Species)) +
  geom_violin() +
  scale_fill_manual(values = c("#69b3a2", "#b37d69", "#b3699f")) +
  xlab("Species")  +
  ylab("Sepal length") +
  ggtitle("Iris flower sepal lengths") +
  theme(legend.position = "none") # Hiding the legend
```

### Time series

```{r time_series}
# Time series in R
data(co2)

plot(co2, col = "#69b3a2", xlab = "year",
    ylab = expression(CO[2] ~ concentration))
```

## Part 2: Tables in R

Tables can be produced in R using `kable()` or you can use a program like
Google Sheets or Microsoft Excel.

```{r kable_table}
# Producing tables using knitR library

tmp <- tempfile(fileext = ".csv")
download.file(url = "https://raw.githubusercontent.com/lundquist-ecology-lab/subway_map_analysis/main/data/subway_data.csv",
            destfile = tmp)
subway_data <- read.csv(tmp, header = TRUE, stringsAsFactors = TRUE)

# Remove "id" and "units" column
subway_data <- subway_data[c(-1, -7)]

# Convert m2 to k2 and round
subway_data$b_area <- round(subway_data$b_area / 1000000, 2)
subway_data$g_area <- round(subway_data$g_area / 1000000, 2)

library(knitr)

# Produce table with more readable heading names and Markdown syntax
kable(subway_data, col.names = c("Borough",
                        "Subway Stations",
                        "Stations near green spaces",
                        "Total borough area (km<sup>2</sup>)",
                        "Total green area (km<sup>2</sup>)"))

# Export table as .csv to work with in other programs

write.csv(subway_data, "subway_table.csv", row.names = FALSE)
```


## Part 3: Figure critique

Based on these rules, we will look at and critique real-world plots. 

* Make sure the data is visible - simple plots are typically better than overly complex ones
* Display patterns in the data clearly
* Be honest, accurately portray the magnitude of changes (i.e. don’t over exaggerate your results)
* Draw your plot clearly, excess stylistic components can hurt instead of help your presentation and use reasonably sized fonts

Plots to review can be found [here](https://www.lundquistecology.com/courses/biostatistics/presenting-data/reviewing-figures).