---
title:  Biol/Math 218 lesson
subtitle: Estimating with uncertainty
author: Your Name Here
output:
  rmdformats::downcute:
    self_contained: true
    thumbnails: true
    lightbox: true
    gallery: false
    highlight: tango
    downcute_theme: "chaos"
    toc_depth: 3
knit: (function(input, ...) {
    rmarkdown::render(
      input)
  })
---

This week we will be discussing sampling and the inherent uncertainty that
comes along with it. We will also be discussing ways to measure uncertainty.

## Part 1: Uncertainty in estimates

**Estimate**: an inference about a population parameter based on a subsample of that
population. Since samples are only a subset of the entire population, there is a 
chance that estimates from repeated sampling from that population can be off by
a particular amount. This uncertainty can be visualized by producing a  
**sampling distribution**. We can model a sampling distribution in R. 


```{r sampling_distribution}

library(palmerpenguins)

data <- na.omit(penguins$bill_length_mm) # Remove blank observations

# There are a total of 342 observations of penguin bill length
# For this, we will assume that there are 344 individuals in the population

pop_mean <- mean(data)
print(pop_mean)

# Random sample of 5
s_5 <- sample(data, size = 5, replace = FALSE)
print(mean(s_5))

# Random sample of 100
s_100 <- sample(data, size = 100, replace = FALSE)
print(mean(s_100))

# Notice that the random sample of 100 is closer to the true population mean.
# Higher sample size (n) results in less uncertainty

# We can visualize this by taking the means of multiple sets of samples
# using a for() loop

# 1000 samples of 5
dis_5 <- NULL
for (i in 1:1000) {
  x <- sample(data, size = 5, replace = FALSE)
  dis_5[i] <- mean(x)
}

# Sampling distributions are reported as frequency histograms
hist(dis_5, col = "#69b3a2", xlab = "Bill length", main = "Samples of 5")


# 1000 samples of 5
dis_100 <- NULL
for (i in 1:1000) {
  x <- sample(data, size = 100, replace = FALSE)
  dis_100[i] <- mean(x)
}

# Sampling distributions are reported as frequency histograms
hist(dis_100, col = "#69b3a2", xlab = "Bill length", main = "Samples of 100")
```


## Part 2: Measuring uncertainty

It is extremely important for uncertainty to be measured and reported for proper 
data analysis and interpretation. There are two ways of measuring uncertainty:

**Standard error**: ± standard deviation of the sampling distribution.

```{r standard_error}
s100 <- sample(data, size = 100, replace = FALSE)

# SE <- sd(y) / sqrt(n)
SE_100 <- sd(s_100) / sqrt(100)
print(SE_100)
# Just like with the sampling distribution, SE will decrease
# (i.e., uncertainty will decrease) as sample size increases.
```

**95% confidence intervals**: the range likely to contain the true population mean

To calculate 95% CI manually, we calculate ȳ ± 2SE

```{r conf_int}
y_bar <- mean(s_100)
print(y_bar)

upper <- y_bar + SE_100
lower <- y_bar - SE_100

print(paste(lower, "±", upper, sep = " "))

# We can round it
print(paste(round(lower, 2), "±", round(upper, 2), sep = " "))
```


# Part 3: Visualizing uncertainty

To visualize uncertainty we use error bars. Error bars can be either standard
deviation, but more commonly either **standard error** or **95% CI**.

**Standard error bars**

We will need to use `ggplot2` to produce a plot with error bars. Here we will
look at the `Palmer Penguin` data bill length data again, but subset it into the
different species of penguin.

```{r penguin_se_plot}
library(palmerpenguins)
data <- penguins
data <- na.omit(penguins)

print(levels(data$species))

library(Rmisc)
# The summarySE() function in the Rmisc library calculates mean, SD, SE, and CI

penguin_summary <- summarySE(data = data, measurevar = "bill_length_mm",
                            groupvar = "species")

print(penguin_summary)

# Bar plot with error bars
library(ggplot2)

y <- penguin_summary$bill_length_mm # set bill length as y for ease

ggplot(penguin_summary, aes(x = species, y = y)) +
  geom_bar(position = position_dodge(0.9), stat = "identity",
          colour = "black", size = .3, fill = "#69b3a2") +
  geom_errorbar(aes(ymin = y - se, ymax = y + se), colour = "black",
        width = .1) +
  ggtitle("Mean bill length ± SE") +
  xlab("Site") +
  ylab("Penguin bill length (mm)") +
  scale_y_continuous(limits = c(0, 60),
                     expand = c(0, 0), # Set y range
                     breaks = 0:100 * 10) +
  theme(panel.grid = element_blank()) +
  theme(legend.position = "none")
```

The code to produce a plot with 95% CI error bars is very similar to ones above, 
but `+ se` and `- se` in the `ggplot()` function is changed to `ci`.

```{r penguin_ci_plot}
library(palmerpenguins)
data <- penguins
data <- na.omit(penguins)

print(levels(data$species))

library(Rmisc)
# The summarySE() function in the Rmisc library calculates mean, SD, SE, and CI

penguin_summary <- summarySE(data = data, measurevar = "bill_length_mm",
                            groupvar = "species")

print(penguin_summary)

# Bar plot with error bars
library(ggplot2)

y <- penguin_summary$bill_length_mm # set bill length as y for ease

ggplot(penguin_summary, aes(x = species, y = y)) +
  geom_bar(position = position_dodge(0.9), stat = "identity",
          colour = "black", size = .3, fill = "#69b3a2") +
  geom_errorbar(aes(ymin = y - ci, ymax = y + ci), colour = "black",
        width = .1) +
  ggtitle("Mean bill length ± 95% CI") +
  xlab("Site") +
  ylab("Penguin bill length (mm)") +
  scale_y_continuous(limits = c(0, 60),
                     expand = c(0, 0), # Set y range
                     breaks = 0:100 * 10) +
  theme(panel.grid = element_blank()) +
  theme(legend.position = "none")
```

Reporting the 95% confidence intervals is particularly useful because in many cases,
if the 95% CI are not overlapping between bars, you can infer that those means
are statistically different from each other. 