```{r introduction}
# Caclulating means in R

# Sample mean (arithmetic mean)

y <- c(5, 3, 2, 6, 4, 3)

n <- length(y)

y_bar <- sum(y) / n

print(y_bar)

# Check using the built-in function
mean(y)

# Sample variance and standard deviation

# Variance

SS <- sum((y - y_bar)^2) # Sum of squares
s_2 <- SS / (n - 1)
# Using n - 1 corrects for bias
print(s_2)

# Check using the built-in function
var(y)

# Standard deviation
s <- sqrt(s_2)
print(s)

# Check using the built-in function
sd(y)
```

### Important: When calculating anything in this course, rounding should only be done at the end. Do not round intermediate calculations.

## Practice problem 1

We will be looking a dataset collected by a fellow Marymount Manhattan student. She
was interested in whether there was an effect of urbanization in different sites
in the [Bronx River](https://en.wikipedia.org/wiki/Bronx_River) on the size of
*Hydropsyche* sp., an important food source for fish in the river.  

[Hydropsyche larva](hydropsyche.png)

In the code block below do the following:

1. Import "body_length.csv" as "b_len"
2. Explore the data using `names` and `head()` and take note of the **column names**
3. Identify the numerical and 
4. Calculate the mean larva body length using the mean equation
5. Check your answer to #4 by using the `mean()` function
6. Calculate the variance and standard deviation of the larva body length using the mean equation
7. Check your answer to #6 by using the `var()` and `sd()` functions 

```{r practice_problems_1}

```


