---
title:  Biol/Math 218 lesson
subtitle: Probability
author: Your Name Here
output:
  rmdformats::downcute:
    self_contained: true
    thumbnails: true
    lightbox: true
    gallery: false
    highlight: tango
    downcute_theme: "chaos"
    toc_depth: 3
knit: (function(input, ...) {
    rmarkdown::render(
      input)
  })
---

This week, we will be working through probability, a core concept in statistics. 

Terms:

- **Probability**: the proportion of random trials that result in a desired event occurs over a number of random trials
- **Mutually exclusive event**: Two events that cannot happen simultaneously 

We will be dealing with probabilities by building probability equations, and then
converting them into syntax that R can recognize. 

**Example probability equation**:

`Pr[flipping heads] = 1/2`

## Part 1: Probability distributions

A probability distribution is the probabilities of all possible outcomes in a 
random trial. 

**Discrete probability distributions** can be produced for discrete numerical events. 

For example the probabilities of rolling three fair six sided dice are:

```{r three_six_sided_dice}
library(dice)

x <- getSumProbs(ndicePerRoll = 3, nsidesPerDie = 6, nkept = 3, dropLowest = FALSE)
x <- data.frame(x[1])

plot(x = x[, 1], y = x[, 2], xlab = "Sum of three dice", pch = 16)
```


**Continuous probability distributions** can be produced for continuous variables. 

For example the probabilities of rolling three fair six sided dice are:

```{r normal_distribution}
y <- rnorm(10000000)

plot(density(y), main = "", xlab = "y", ylab = "Probability density")
```

## Part 2: Addition rules

If two events A and B are mutually exclusive

`Pr[A OR B] = Pr[A] + Pr[B]`

and the probability of an event not occurring is

`Pr[not A] = 1 - Pr[A]`

If A and B are not mutually exclusive we use the **general addition rule**:

`Pr[A OR B] = Pr[A] + Pr[B] - Pr[A and B]`

## Part 3: Multiplication rules

If two events A and B are independent,

`Pr[A AND B] = Pr[A] * Pr[B]`

However, if events are dependent on one another, the 
probabilities become a little more complicated. 

We can visualize the probability of a sequence of dependent (or independent)
events happening using probability trees.

### Example: Probability trees

Biological sex in turtles at birth is determined by 
the temperature at which the eggs were incubated at. Turtle offspring
incubated at 27.7 °C are 100% male, and those incubated above 31 °C are
100% female. However, if the temperatures fluctuate between those two extremes, 
there will be a mixture male and female hatchlings. 

Let's say that under normal circumstances with fluctuating temperatures, there 
is a 50% chance of a hatchling being female. However, if there are more 
days closer to 31 °C, the probability is instead 80% of a hatchling being 
female or male.

Let's also suppose that under normal circumstances, there is a 15% chance of the 
more days being above 31 °C and there is no chance of more days at or below 27.7 °C.

What is the probability that a hatchling will be male under "normal" conditions?

```{tree}
                    Majority of days  |-------0.8--------> Female
                    are above 31 °C   |
                                      |
    |---------0.15----------> Yes   --|    
    |                                 |
    |                                 |
    |                                 |-------0.2--------> Male
----|                                     
    |                                 |-------0.5--------> Female
    |                                 |
    |                                 |
    |--------0.85-----------> No    --|
                                      |
                                      |
                                      |-------0.5-------> Male
```    

`Pr[hatchling is male] = Pr[male AND above 31] + Pr[male AND fluctuating]`

```{r probability}
p_hatch_m <- (0.15 * 0.2) + (0.85 * 0.5)
print(p_hatch_m)
```

### Practice problem 1: Effect of global warming on turtle sex ratios

It is important to consider the biological impacts of global changes and how
they might impact different species. For example, the mean global temperature is
increasing year after year and this could have a major impact on species with
temperature-dependent sex determination like turtles. Lets say that in the next 
100 years, the probability of the majority of days being above 31 °C increases to
30%, how will that impact the sex ratio of turtle populations? 

Below, fill out a new probability tree based on the new and calculate the probability that a 
hatchling will be female. 

```{new_tree}
```

`New equation`

```{r new_prob}
```

What do you think would be the **biological significance** of this sort of global change?


## Part 4: Conditional probabilities and Bayes' theorem

**Conditional probability** is the probability that an event occurs given that
another event occurs

`Pr[A|B]`

Using conditional probabilities, we can calculate the 
overall probability of an event by invoking the **law of total probability**

`Pr[hatchling female] = Pr[temps above 31 °C] * Pr[female|temps above 31 °C] + Pr[temps fluctuate] * Pr[female|fluctuate]`

We can go further and determine the probability of two events happening, even if 
they are dependent by using the *general **multiplaction rule**

`Pr[A and B] = Pr[A] * Pr[B|A]`

Note, it does not matter which event we label A or B.

**Bayes' theorem** is a special use of conditional probabilities where

`Pr[A|B] = (Pr[B|A] * Pr[A]) / Pr[B]`

Bayes' theorem has a lot of use in the medical field. For example, it can be 
used to detect false positives in medical diagnoses. 

### Practice problem 2: Bayes' theorem

Lets imagine a hypothetical deadly disease that we have a test to detect. The rate of
correctly detecting a patient with this disease is  70%. We can call this the 
**sensitivity** of the test. However, the rate at which patients without the disease
are misdiagnosed in 4%. We call this a **false positive**. This sounds pretty good,
but what is the probability that a patient that tests positive for the disease 
actually has the disease?

There are a lot of conditional probabilities here, so let's express them using 
Bayes' theorem:

`Pr[disease|positive] = (Pr[positive|disease] * Pr[disease]) / Pr[positive]`

We will work through this together in class

```{r Bayes_theorem}
```