---
title:  Biol/Math 218 lesson
subtitle: Non-parametric tests
author: Your Name Here
output:
  rmdformats::downcute:
    self_contained: true
    thumbnails: true
    lightbox: true
    gallery: false
    highlight: tango
    downcute_theme: "chaos"
    toc_depth: 3
knit: (function(input, ...) {
    rmarkdown::render(
      input)
  })
---

Up to this point, we have been testing whether the assumptions of our particular
test has been met by our data. However, it is not guaranteed that our assumptions
will be met in all cases. If we find that the assumptions have been violated, we
have a few options:

## Option 1: Ignore the violations
Ignore the violations: This usually only works when we are comparing means and 
the normality assumption is violated. If our *n* is high, we can 
ignore slight non-normality. We can't really do this for differences in variance,
especially if our samples are small or our uneven.

## Option 2: Transform the data
We can use different transformations on the data and then retest the assumptions.
There are a wide variety of transformations we can do.

Here are some examples:

```{r data_transformations}
# Raw data
data <- iris$Sepal.Width
hist(data)

# Natural log transformation
log_transform <- log(data + 1)
hist(log_transform)

# Log_10 transformation
log10_transform <- log10(data + 1)
hist(log10_transform)

# Square-root transformation
sqrt_transform <- sqrt(data + 0.5)
hist(sqrt_transform)
```

Important notes on transformations:

1. While analyses should be done using **transformed** data, data should be presented and reported **untransformed**
2. All data used in a particular study needs to be transformed using the same technique.

## Option 3: Use nonparametric tests

If transformation does not work, **nonparametric tests**, which do not rely on
population parameters, can be used instead. There are a number of analogous 
nonparametric tests we can use in place of parametric tests.

### Sign test

Replaces one-sample t-test or paired t-test. Basically, results above the 
median are assigned a "+" and those below are assigned a "-". It is very similar
to a binomial test. Because a lot of information from the data is not used, it 
has a lower power than those parametric tests.

Assumptions:

- All samples are random samples

```{r sign_test}
# One sample
data(iris)
sample <- iris$Petal.Width[iris$Species == "versicolor"]

# Hypothetical median (median of all Iris species' petal widths)
med <- median(iris$Petal.Width)

signs <- NULL
for (i in 1:length(sample)) {
  if (sample[i] > med) {
    signs[i] <- paste("+")
  } else {
    signs[i] <- paste("-")
  }
}
print(signs)

trials <- length(sample)
print(trials)

successes <- length(signs[signs == "+"])
print(successes)

# Sign test
binom.test(successes, trials)

# OR

library(rstatix)
sample <- data.frame(sample)
sample %>% sign_test(sample ~ 1, mu = med)

# Paired data

before <- c(450, 400, 440, 390, 400, 432, 415, 401)
after <- c(510, 590, 520, 550, 585, 500, 590, 577)

data <- data.frame(weight = c(before, after),
                  timing = c(rep("before", 8), rep("after", 8)))

# Sign test
data %>% sign_test(weight ~ timing)
```


### Mann-Whitney U-test or Wilcoxon rank sum test

This replaces the two-sample t-test. This test uses the **ranks** (smallest to
largest) of the data instead of the absolute values.

Assumptions:

- All samples are random samples
- The shape of the distribution of the measurement have the same shape

```{r mann_whitney_test}
library(palmerpenguins)
data(penguins)

penguins <- na.omit(penguins)

adelie_bill <- penguins$bill_length_mm[penguins$species == "Adelie"]
gentoo_bill <- penguins$bill_length_mm[penguins$species == "Gentoo"]

# Mann-Whitney U-test/Wilcoxon rank sum
wilcox.test(adelie_bill, gentoo_bill)
```

### Kruskal-Wallis test

This test replaces the ANOVA. This test is based on ranks, similarly to the 
Wicoxon rank sum test and Mann-Whitney U-test

Assumptions:

- All samples are random samples
- The shape of the distribution of the measurement must be the same for all populations

```{r kruskal_wallis_test}
library(palmerpenguins)
data(penguins)

penguins <- na.omit(penguins)

# Kruskal-Wallis test
kruskal.test(bill_length_mm ~ species, data = penguins)
```

