---
title:  Biol/Math 218 lesson
subtitle: Introduction and basic data management
author: Your Name Here
output:
  rmdformats::downcute:
    self_contained: true
    thumbnails: true
    lightbox: true
    gallery: false
    highlight: tango
    downcute_theme: "chaos"
    toc_depth: 3
knit: (function(input, ...) {
    rmarkdown::render(
      input)
  })
---

This week we will go over the syllabus and learn how to download and use code
off of the [GitHub repository](https://github.com/lundquist-ecology-lab/biostatistics)
and how to use RMarkdown to write R notebooks and produce and submit 
portfolio assignments.

## Part 1: Using GitHub

If you are reading this in RStudio on your 
[personal RStudio instance](https://lundquistecology.com/getting-to-know-your-rstudio-server-instance/), 
you are read to go on to **Part 2**.

## Part 2: Using R Markdown in RStudio

R Markdown files (`.Rmd`) are a combination of `R` (or `Python`) code blocks and
the writeup language called `Markdown`, which is itself based on `HTML` which is
the basic language of webpages. Using R Markdown allows us to write in regular prose
as well as write executable code in code blocks. It also allows us to export 
that prose and code as a familiar format like `.docx` (Microsoft Word), `.pdf` (Adobe), 
and `.html` (website). In our case, you will be exporting your portfolios as 
quite stylish `.html`.

Check out this [page](https://rmarkdown.rstudio.com/articles_intro.html) for a 
run down of anatomy of a R Markdown document as well
as a step by step guide to export (or **knit**) a `.Rmd` file to `.html`.

## Part 3: Dealing with data

Data is made up of **variables** can classify data into two basic:

- **Numeric**: numbers, either whole or fractions
    - Discrete: Only certain values possible within a given range (e.g. counts)
    - Continuous: Any value is possible within a given range (e.g. measurements)

- **Categorical** (factors): identifies a grouping of data

Both numeric is collected through **experimental** and **observational** studies
and most data is recorded electronically as either Excel spreadsheets (`.xlsx`) 
or as comma separated value sheets (`.csv`) where all values of are separated 
from each other with a `,`. 

Example csv format:

```{r csv_example, eval = FALSE}
variable_1, variable_2, group
10, 15, 1
11, 13, 1
12, 11, 1
14, 11, 1
12, 11, 2
11, 11, 2
15, 13, 2
16, 12, 2
```
Notice the first row contains a variable name for each colum and then each other row
is a set of measurements (numeric) or identifiers (factors). 

- Each row represents a single observation of a particular set of variables. 
- Each column represents all observations of a particular variable.

## Important concepts
- The entire table together is a `data frame`
- Each column represents a single `vector` or `array` of data
- Data in columns can represent `numerical` data or made up of `factors`, even if the arrays are made up of numbers
- We need to explicitly indicate which variables are `factors` in our code

### Exercise: Importing data

Built-in data:
```{r built_in_data}

# Importing built in data

data(iris) # This imports the built-in Iris dataset

print(iris$Species) # Example code, printing the species of Iris in the dataset

# List all built-in datasets

data()
```

.csv files (from the internet)

```{r csv_internet}
tmp <- tempfile(fileext = ".csv")
download.file(url = "https://data.cdc.gov/resource/9bhg-hcku.csv",
            destfile = tmp)
covid_data <- read.csv(tmp, header = TRUE, stringsAsFactors = TRUE)
head(covid_data) # Gives us the first few rows of the data
```

.xlsx files (from the internet)

```{r xlxs_internet}
library("readxl")
tmp <- tempfile(fileext = ".xlsx")
download.file(url = "https://dryad.org/stash/downloads/file-stream/1936219",
            destfile = tmp)
brassica_data <- read_excel(tmp)
head(brassica_data)
```

If the files are on local storage

```{r local_files}

# If the data is in .csv format
DATA_FILE_CSV <- read.csv("/path/to/DATA_FILE.csv", header = TRUE, stringsAsFactors = TRUE)

# If the data is in .xlxs format
library("readxl")
DATA_FILE_XLXS <- read_excel("/path/to/DATA_FILE_XLXS")
```

There may be other data file formats out there (e.g. .txt), and you can search
on Google how to import those.

## Exercise exploring data

We can explore the data using functions in R

```{r explore_data}
# Explore built-in data
data()

# Download data from an R library()
library(palmerpenguins) # Data is in a library called "palmerpenguins"
palmerpenguin <- data(penguins)

# Some diagnostic functions

# First few rows
head(penguins)
# Last few rows
tail(penguins)

# Determine if particular variable is a factor or numeric
class(penguins$species)
class(penguins$bill_length_mm)
```